\chapter{Finetuning TabPFN-TS}
TabPFN-TS already showed remarkable results on the GIFT-Eval benchmark.
But as shown in Fig 2 TODO it still especially struggles with noisy and non-periodic time series.
\newline
In this Section we will further investigate the potential benefits of finetuning TabPFN-TS with the time series datasets of GIFT-Eval.
For this, we will look at two different ways of finetuning.
The first is finetuning on one dataset.
This will presumably enhance performance on this specific dataset and potentially other datasets that are the same. %todo
The second method is finetuning on multiple datasets.
As \citep{Hollmann_2025} already showed, finetuning TabPFN on multiple datasets acts just like continuing pretraining so it enhances performance on a wide range of different dataset.
We will further investigate this by applying this method to time series data.

\section{Method}
- finetuning noch mal genauer erklären
We will use the full finetuning methode where we update all parameters of the model.
This method has been proved to be the best way of finetuning TabPFN since it is tolerant to full updates \citet{tanna2026exploring}.


- besonderheiten die geändert wurden von den stand von lennart purucker
- daten erstellung
- validation
- sagen welche datensets wir nehmen für finetunen -> alle außer welche mit variabler Ts länge weil da padding in spiel kommt und das erstmal nicht angeschaut wird
- hyperparameter for every dataset
- l2 sp norm from realtabpfn
- context length max 4096 from gift eval config
- validation configs
- bisschen an real tabpfn orientieren
- nur short prediction gefinetuned wegen technical limitations
\section{Experiments}
\subsection{Finetuning on one Dataset}
- auch schauen wie performance auf einem datenset aussieht nach dem auf einem anderen gefinetuned wurde
- top 5 hp-konfigurationen mit bestem val loss werden genommen
- graph zeigen
-
\subsection{Finetuning on multiple Datasets}