\chapter{Introduction}

\section{Motivation}
Since the release of ChatGPT in 2022 and the widespread media coverage of AI, machine learning and deep learning have become more accessible to the general public.
Large Language Models are now known as being a gateway to machine learning and artificial intelligence.
But LLMs almost only excel on data in textual format or pictures.
What about tabular data?
Wether it is patients data in healthcare, sales data in finance or every other data which can be stored in an Excel-like schema, tabular data is one of the most common data formats used in a big scale.
This type of data has not been in the focus by researchers for quite some time since nobody found a way to utilize deep learning for tabular data.
\newline
Furthermore, algorithms like gradient boost decision trees or random forest were much more accessible.
But with the release of TabPFN, a transformer model pre-trained to solve artificially generated classification tasks from a synthetic prior, a new benchmark for tabular inference tasks has been set.
In a single forward pass, TabPFN is able to solve classification, regression, and inference tasks.
%\\
%Early 2025 TabPFN-TS was released.
%A method that maps time series data onto tabular data to achieve SOTA forecasting results with the already pre-trained TabPFN model.
\\
There are two main ways to presumably enhance the performance of this specific model.
One is finetuning a pre-trained model and the other is continued pre-training a pre-trained model.
The key difference between those two is the data the model will further be trained with.
Only by using real-world datasets for a specific domain like healthcare etc. finetuning trains a pre-trained model to enhance the accuracy of it in this domain. %TODO
In continued pre-training one uses a specific synthetic prior to generate synthetical classification, regression, or inference tasks for the domain to further train and enhance the model.
\\
Although there are already some models which achieve top rank results in time series forecasting, the impact of finetuning or continued pretraining has not been studied and potential benefits remain uncovered.

\section{Exposé}
TabPFNv2 can work with time-series as well, as \citet{hoo2025tables} showed, although it has not been trained with time-series data during pre-training, treating time series forecasting as a tabular regression problem.
Good performance in time-series data derives from two main causes: Preprocessing the timestamp feature of the data into Automatic Seasonal Features, Calender Features and Running Index and a pretrained model which has a strong zero shot performance on regression tasks. \newline
This thesis focuses on finetuning and continuing pretraining on TabPFNv2 with time series data.
We will address the concept of the aforementioned methods and examine wether or not these methods lead to performance boosts in time series forcasting methods.
\newline
The questions we will address are:
\begin{enumerate}
    \item Can we finetune TabPFN-TS on time-series data (and does it improve performance)?
    \item Can we further improve performance by finetuning on multiple datasets similar to RealTabPFN \citep{Hollmann_2025}
    \item Can we build a prior that generates realistic time-series data?
    \item Can we continue pre-training TabPFN-TS with a given prior?
    \item Which one of the following models performs best on various high scale benchmarks: TabPFN-TS, finetuned TabPFN-TS, or continued pre-trained TabPFN-TS?
\end{enumerate}
Chapter 4 will focus on finetuning the model with datasets from the GIFT-EVAL benchmark as well as datasets from autogluon/chronos-datasets.
Most of the code used in this section was already provided by Lennart Purucker, but we had to adjust some of the code to bring it up to date and to use it with time series data.
To examine the presumable improvements we use the GIFT-EVAL evaluation code to calculate different types of errors before and after finetuning.
\newline
Chapter 5 covers the process of continued pretraining TabPFNv2 on a given prior.

\section{Related Work}
\subsection{TabPFN}
Released by the University of Freiburg TabPFN is a prior fitted network which builds up on a 12 layer transformer.
It is able to solve tabular classification and regression tasks in a single forward pass.
\newline
Contrary to the standard supervised deep learning approach by which models are trained from scratch for every dataset TabPFN is pretrained once by a large amount of artificial training tasks.
Normally in machine learning, a training task consist of a tuple with the following scheme $(x_{train}, y_{train})$ in which $x_{train}$ represents the input data and $y_{train}$ the corresponding labels.
But with TabPFN it is a bit different since a tabular data consists of more than one tuple.
So one datapoint which is used to train TabPFN is a set of training samples $D_{train} = {(x_1, y_1), ..., (x_n, y_n)}$.
The key problem with this type of training is that tabular datasets with this size are extremely rare and often not accessible without breaking privacy laws.
To obtain a sufficient amount of data to train TabPFN with, a prior is used.
\newline
A prior is a generative process that samples artificial tabular based on different hyperparameters like length of dataset, amount of features etc.
To be more precise, TabPFN uses Structural Causal Models(SCM) which are based on graphs to generate this data(More in Preliminaries).
By sampling different characteristics of the underlying graph from the prior's distribution a very heterogeneous set of datasets can be sampled.
For the whole training process synthetic datasets $D := (x_i, y_i)_{i \in \{1, ..., n\}}$ are sampled and used to optimize the parameters of TabPFN.
To obtain the best result, the prior has to cover a heterogeneous set of datasets to be able to learn a generic algorithm.
\newline
TabPFNs architecture is a combination of the original transformer encoder and the original PFN architecture.
As said earlier TabPFN uses z normalization per feature and maps all numerical values to floating points and categorical values to integers
The most import part in the architecture is that it uses inter-feature and sample attention.
But unlike fully connected attention, test samples are not allowed to attend each other and only to the training data, so that test samples influence each other.
\newline
TabPFN training and inference differs fundamentally from traditional deep learning algorithms.
As previously mentioned, for TabPFN one training sample consists of a set of samples ...%TODO
Instead of doing one forward pass for each input data point, at inference time the pretrained TabPFN model will receive $D_{train} = {(x_1, y_1), ..., (x_n, y_n)}$ for context and predicts $y_{test_pred}$ based on $X_{test}$ in one single forward pass.
So it does fitting and prediction in one forward pass.
To generalize prediction over the vast amount of different kinds tabular data can occur TabPFN uses artificially generated training data generated from one or respectively multiple priors
and so approximates the posterior predictive distribution
\[
    p(\hat{y}|x_{test}, D_{train}) \propto \\int_{\Phi} p(\hat{y}|x_{train}, \phi) p(D_{train}|\phi) p(\phi) d\phi
\]
where $D_{train}$ is the context, $x_{test}$ are the labels and $\Phi$ is the set of priors used for pretraining.
Instead of returning a single point estimation in regression, TabPFN returns a target distribution, capturing the uncertainty of predictions.
The borders of this predictive distribution are calculated by choosing the 1/5000 quantiles of the prior distribution which TabPFN was pretrained on.
They are normalized since TabPFN was trained on z-normalized data in pretraining but will be reverted by applying the inverse of the transform to the borders of the buckets.
For training for regression, the whole distribution is used to calculate the minimize the mean negative log density, whereas for inference mostly the median of the probabilistic distribution is used.
%classification point estimation?

\newline

- weitere fakten
- excels on small sample datasets <10000 500 features
- no hyperparamter tuning
- better than other methods like XGBoost
- because of architecture invarian
\subsection{Time Series Forecasting}
- time series forecasting really relevant(?)
- pfn achieve top rank results
\subsubsection{TabPFN-TS}
\citet{hoo2025tables} introduces TabPFN-TS: A model that combines TabPFN-v2 with lightweight feature engineering to enable both point and probabilistic forecasting.
- released by Uni freiburg
- is a wrapper for Tabpfn
- uses feature transformation for time series data
- time series data is nothing else than tabular data
- achieves top rank results in time series forecasting
\subsubsection{Other work}
- ForecastPFN(from prior)
- %TODO search for SOTA time series forecasting methods
\subsection{Prior Building}
- \citet{dooley2023forecastpfn} is especially relevant since they created a synthetic prior for time-series data.
They create the synthetic data by generating a trend, seasonal, and noise factor all either observed once a day, month, or year.
\subsection{Finetuning}
Continued pre-training on large real-world datasets in a specific domain yields superior downstream predictive accuracy in the specified domain compared to an only pre-trained model \citep{Hollmann_2025} .
This validates the assumption that TabPFN performs better if we adapt weights to a downstream task.\newline
%TabTune
%on finetuning tabular models
\subsection{Continued Pretraining}
In \citet{kolberg2025tabpfn} we saw that if we continue pre-training TabPFNv2 (so continued training on a synthetic prior) in a similar modality (tabular data with high feature and low sample count) will not decrease performance on non high dimension, low sample size data.\newline

%https://arxiv.org/abs/2506.08982
%

%Tabpfn
%time series forecasting
    %Tabpfn ts
    %other works
%prior building
%Finetuning
%continued pretraining
